---
title: "PCaMM Data Cleaning - Chemistry"
output: html_document
---

This is the data cleaning script for the sediment and water chemistry for the PCaMM project.
This cleans only the data for the experimental work, and does not include the data for the preliminary studies.

First we'll get our environment set up:

```{r set_up_environment, message=FALSE, warning=FALSE, results='hide'} 
rm(list=ls())
library(openxlsx)
library(readxl)
library(tidyverse)
source("prepare_metadata.R")
rm(PW_container_info, sediment_container_info)
```

# Set up for PW water chemistry data

First, we'll prepare the water chemistry data that we have for the porewater flowing through the sediment cores.
To do this, we'll start with making the dataframe:

```{r prep_dataframe, message=FALSE, warning=FALSE, results='hide'} 
water_chemistry_data <- incubation_metadata
PW_container_to_inc_key <- PW_metadata %>%
  select(containerID, incubationID)
```

## Read in and prepare DOC data

We'll then add the DOC data, cleaning up some of the column names and making the DOC values numeric.
This data was pulled from the database on 2026-03-13.

```{r read_in_DOC_data, message=FALSE, warning=FALSE, results='hide'}
doc_data_raw <- read_excel("../dataRaw/20260313_PCaMM_DOC_databasepull.xlsx") %>%
  rename_with(~ gsub(" ", "_", .x)) %>%
  rename(containerID = DOC_Data.ContainerID) %>%
  mutate(DOC_mgC_L = as.numeric(DOC_mgC_L)) %>%
  filter(!is.na(DOC_mgC_L))
```

Next we'll want to add the incubationID to the DOC data, so that we can join it to the water chemistry data.
Then, clean up.

```{r clean_up_DOC_data, message=FALSE, warning=FALSE, results='hide'}
# Add incubation ID
doc_data <- PW_container_to_inc_key %>%
  right_join(doc_data_raw,
            by = "containerID") %>%
  filter(!is.na(incubationID)) %>%
  select(-containerID)
# Add to water chemistry data
water_chemistry_data <- water_chemistry_data %>%
  left_join(doc_data,
            by = "incubationID")
# Clean up
rm(doc_data_raw, doc_data)
```

## Read in and prepare sulfide data

Next add the sulfide data.
This data was not pulled from the database, but rather is stored in the original data frames generated on 2024-07-11, 2024-07-17, and 2024-08-02.

Since this is a little more comple, I'm going to make a function to read in the sulfide data, and then use that function to read in the three data files and combine them into one dataframe.

```{r function_to_read_in_sulfide_data, message=FALSE, warning=FALSE, results='hide' }
# Function to read in sulfide data
read_sulfide <- function(datafile) {
  read_excel(datafile,
             sheet = "Sample Measurement") %>%
    rename_with(~ gsub(" ", "_", .x)) %>%
    rename(sulfide_ppm = `Original_[S(-II)]_(mg/L)`,
           containerID = Sample_ID,
           sulfide_ppm_QA = QA_Flag) %>%
    select(containerID, sulfide_ppm, sulfide_ppm_QA) %>%
    filter(!is.na(containerID))
}
```

Now call that function, add the incubationID, and join to the water chemistry data.

```{r read_in_sulfide_data, message=FALSE, warning=FALSE, results='hide'}
# Read in and combine sulfide data
sulfide_data <- rbind(read_sulfide("../dataRaw/sulfide/20240711_Sulfide_E1.xlsx"),
                      read_sulfide("../dataRaw/sulfide/20240717_Sulfide_E1.xlsx"),
                      read_sulfide("../dataRaw/sulfide/20240802_Sulfide.xlsx")) %>%
  mutate(sulfide_ppm = as.numeric(sulfide_ppm))
# Add incubationID to sulfide
sulfide_data <- PW_container_to_inc_key %>%
  right_join(sulfide_data) %>%
  filter(!is.na(incubationID)) %>%
  select(incubationID, sulfide_ppm, sulfide_ppm_QA)
# Add to water chemistry data
water_chemistry_data <- water_chemistry_data %>%
  left_join(sulfide_data,
            by = "incubationID")
rm(sulfide_data, read_sulfide)
```

## Read in and prepare porewater MeHg data

Next read in the porewater MeHg data.
This data is stored in the file `PCaMM_PW_MeHg.xlsx`.

```{r read_in_MeHg_data, message=FALSE, warning=FALSE, results='hide' }
mehg_data <- read_excel("../dataRaw/hg_pw/PCaMM_PW_MeHg.xlsx") %>%
  rename(PW_MeHg_amb_ppt = `Ambient MeHg`,
         PW_MeHg_201_ppt = `MeHg from Source_1 (201)`,
         PW_MeHg_204_ppt = `MeHg from Source_2 (204)`
  ) %>%
  select(containerID, PW_MeHg_amb_ppt, PW_MeHg_201_ppt, PW_MeHg_204_ppt) %>%
  filter(!is.na(containerID)) %>%
  mutate(PW_MeHg_amb_ppt = as.numeric(PW_MeHg_amb_ppt),
         PW_MeHg_201_ppt = as.numeric(PW_MeHg_201_ppt),
         PW_MeHg_204_ppt = as.numeric(PW_MeHg_204_ppt))
```

Now add the incubationID, and join to the water chemistry data.

```{r add_incubationID_to_MeHg_data, message=FALSE, warning=FALSE, results='hide'}
# Add incubationID to MeHg
mehg_data <- PW_container_to_inc_key %>%
  right_join(mehg_data) %>%
  filter(!is.na(incubationID)) %>%
  select(incubationID, PW_MeHg_amb_ppt, PW_MeHg_201_ppt, PW_MeHg_204_ppt)
# Add to water chemistry data
water_chemistry_data <- water_chemistry_data %>%
  left_join(mehg_data,
            by = "incubationID")
rm(mehg_data)
# Clean up water chemistry data
rm(PW_container_to_inc_key, PW_metadata,
   incubation_metadata)
```

# Sediment Hg data preparation

Next, we'll start on the sediment Hg data, which is the only measurements we have on the solid fraction.
First we'll prep the dataframe:

```{r prep_sediment_Hg_data, message=FALSE, warning=FALSE, results='hide'}
sediment_Hg_data <- sediment_metadata
sed_container_to_inc_key <- sediment_metadata %>%
  select(containerID, incubationID)
```

## Prepare sediment MeHg data

Next, read in and prepare sediment MeHg data:

```{r read_in_sediment_MeHg_data, message=FALSE, warning=FALSE, results='hide'}
temp_mehg_data <- read_excel("../dataRaw/20260728_PCaMM_ExperimentalSeds_MeHg_THg_Results.xlsx",
                             sheet = "MeHg") %>%
  rename(containerID = `Container ID`,
         sed_MeHg_amb_ng_g = Ambient_MeHg_conc,
         sed_MeHg_amb_flag = Ambient_MeHg_flag,
         sed_MeHg_201_ng_g = `201_MeHg_conc`,
         sed_MeHg_201_flag = `Source1_MeHg_flag`,
         sed_MeHg_204_ng_g = `204_MeHg_conc`,
         sed_MeHg_204_flag = `Source2_MeHg_flag`,) %>%
  select(containerID, sed_MeHg_amb_ng_g, sed_MeHg_amb_flag,
         sed_MeHg_201_ng_g, sed_MeHg_201_flag,
         sed_MeHg_204_ng_g, sed_MeHg_204_flag)
```

Now, let's just see what the distribution of data looks like, both as summary statistics:

```{r summary_of_sediment_MeHg_data, message=FALSE, warning=FALSE, echo=FALSE}
summary(temp_mehg_data %>% select(-containerID, -sed_MeHg_amb_flag, -sed_MeHg_201_flag, -sed_MeHg_204_flag))
```

Also visualize as a boxplot:
 
```{r plot_distribution_of_sediment_MeHg_data, message=FALSE, warning=FALSE, echo=FALSE, fig.width=5, fig.height=4}
temp_mehg_data %>%
  select(-sed_MeHg_amb_flag, -sed_MeHg_201_flag, -sed_MeHg_204_flag) %>%
  pivot_longer(cols = -containerID,
               names_to = "MeHg_type",
               values_to = "MeHg_conc") %>%
  filter(!is.na(MeHg_conc)) %>%
  ggplot(aes(x = MeHg_type, y = MeHg_conc)) +
  geom_boxplot() +
  geom_jitter(width = 0.1) +
  scale_y_log10() +
  labs(title = "Distribution of sediment MeHg data",
       x = "MeHg type",
       y = "MeHg concentration (ng/g, log scale)") +
  # Make the x-axis labels more readable
  scale_x_discrete(labels = c("sed_MeHg_amb_ng_g" = "Ambient",
                              "sed_MeHg_201_ng_g" = "Source 1 (201)",
                              "sed_MeHg_204_ng_g" = "Source 2 (204)")) +
  theme_bw()
```

I don't see any immediate causes for concern, so let's join it to the sediment Hg data:

```{r joining_MeHg_to_sediment_data, message=FALSE, warning=FALSE}
sediment_Hg_data <- sediment_Hg_data %>%
  left_join(temp_mehg_data,
            by = "containerID")
rm(temp_mehg_data)
```

## Prepare sediment THg data

Next, read in and prepare sediment THg data:

```{r read_in_sediment_THg_data, message=FALSE, warning=FALSE, results='hide'}
temp_thg_data <- read_excel("../dataRaw/20260728_PCaMM_ExperimentalSeds_MeHg_THg_Results.xlsx",
                            sheet = "THg") %>%
  rename(containerID = `Container ID`,
         sed_THg_amb_ng_g = Ambient_HgT_conc,
         sed_THg_amb_flag = Ambient_HgT_flag,
         sed_THg_201_ng_g = `201_HgT_conc`,
         sed_THg_201_flag = `201_HgT_flag`,
         sed_THg_204_ng_g = `204_HgT_conc`,
         sed_THg_204_flag = `204_HgT_flag`,) %>%
  select(containerID, sed_THg_amb_ng_g, sed_THg_amb_flag,
         sed_THg_201_ng_g, sed_THg_201_flag,
         sed_THg_204_ng_g, sed_THg_204_flag)
```

Now, let's just see what the distribution of data looks like, both as summary statistics:

```{r summary_of_sediment_THg_data, message=FALSE, warning=FALSE, echo=FALSE}
summary(temp_thg_data %>% select(-containerID, -sed_THg_amb_flag, -sed_THg_201_flag, -sed_THg_204_flag))
```

Also visualize as a boxplot:
 
```{r plot_distribution_of_sediment_THg_data, message=FALSE, warning=FALSE, echo=FALSE, fig.width=5, fig.height=4}
temp_thg_data %>%
  select(-sed_THg_amb_flag, -sed_THg_201_flag, -sed_THg_204_flag) %>%
  pivot_longer(cols = -containerID,
               names_to = "THg_type",
               values_to = "THg_conc") %>%
  filter(!is.na(THg_conc)) %>%
  ggplot(aes(x = THg_type, y = THg_conc)) +
  geom_boxplot() +
  geom_jitter(width = 0.1) +
  scale_y_log10() +
  labs(title = "Distribution of sediment THg data",
       x = "THg type",
       y = "THg concentration (ng/g, log scale)") +
  # Make the x-axis labels more readable
  scale_x_discrete(labels = c("sed_THg_amb_ng_g" = "Ambient",
                              "sed_THg_201_ng_g" = "Source 1 (201)",
                              "sed_THg_204_ng_g" = "Source 2 (204)")) +
  theme_bw()
```

I don't see any immediate causes for concern, so let's join it to the sediment Hg data.
Clean up the rest as well.

```{r joing_THg_to_sediment_data, message=FALSE, warning=FALSE}
sediment_Hg_data <- sediment_Hg_data %>%
  left_join(temp_thg_data,
            by = "containerID")
rm(temp_thg_data, sed_container_to_inc_key, sediment_metadata)
```


## Calculate % MeHg

For these analyses, our main interest is in the % MeHg of the total Hg in the sediment.

```{r calculate_percent_MeHg, message=FALSE, warning=FALSE}
sediment_Hg_data <- sediment_Hg_data %>%
  mutate(sed_percentMeHg_amb = sed_MeHg_amb_ng_g / sed_THg_amb_ng_g * 100,
         sed_percentMeHg_201 = sed_MeHg_201_ng_g / sed_THg_201_ng_g * 100,
         sed_percentMeHg_204 = sed_MeHg_204_ng_g / sed_THg_204_ng_g * 100)
```

Let's check on distribution of % MeHg, of course:

```{r plot_distribution_of_percent_MeHg_data, message=FALSE, warning=FALSE, echo=FALSE, fig.width=5, fig.height=4}
sediment_Hg_data %>%
  select(sed_percentMeHg_amb, sed_percentMeHg_201, sed_percentMeHg_204) %>%
  pivot_longer(cols = everything(),
               names_to = "percentMeHg_type",
               values_to = "percentMeHg") %>%
  filter(!is.na(percentMeHg)) %>%
  ggplot(aes(x = percentMeHg_type, y = percentMeHg)) +
  geom_boxplot() +
  geom_jitter(width = 0.1) +
  scale_y_log10() +
  labs(title = "Distribution of sediment % MeHg data",
       x = "% MeHg type",
       y = "% MeHg (log scale)") +
  # Make the x-axis labels more readable
  scale_x_discrete(labels = c("sed_percentMeHg_amb" = "Ambient",
                              "sed_percentMeHg_201" = "Source 1 (201)",
                              "sed_percentMeHg_204" = "Source 2 (204)")) +
  theme_bw()
```

----

# Save out data

Save out all data from this cleaning process, into two separate sheets in an Excel file.

```{r save_out_data, message=FALSE, warning=FALSE}
data_to_save <- list("Water_chemistry_data" = water_chemistry_data,
                     "Sediment_Hg_data" = sediment_Hg_data)
write.xlsx(data_to_save,
           file = "../dataCleaned/incubation_chemistry_data.xlsx",
           asTable = TRUE)
```

Also save it as an RData file, so that we can load it in later without having to read in the Excel file.

```{r save_out_RData, message=FALSE, warning=FALSE}
save(water_chemistry_data, sediment_Hg_data,
     file = "../dataCleaned/incubation_chemistry_data.RData")
```

